---
title: "DATA602WineProject"
author: "Yu Hao"
date: "2025-10-06"
output: pdf_document
---

``` {r, message=FALSE, warning=FALSE}
rm(list=ls(all=TRUE))
library(ggplot2)
library(mosaic)
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

## Question 2: Is the pH level of the resultant *vinho verde* red wine significantly correlated with its perceieved quality?

### Statistical Hypotheses
$H_0$: The pH level of the red wine does not have a significant linear relationship with its quality score.

$H_A$: The pH level of the red wine has a significant linear relationship with its quality score. 

$$
\begin{aligned}
H_0&:\beta_1=0\\
H_A&:\beta_1\neq0\\
\end{aligned}
$$

#### Assumption Checking
Let $X$ represent the pH level of a bottle of *vinho verde* red wine and let $Y$ represent the quality score of a bottle of *vinho verde* red wine. 

Relationship between $X$ and $Y$ variables
``` {r 2.a1}
wine<-read.csv("C:\\Users\\yuhao\\Desktop\\Uni\\DATA 602\\winequality-red.csv")

ggplot(wine, aes(x=pH, y=quality))+
  geom_point(col="blue")+
  xlab("pH Value")+
  ylab("Quality Score")+
  ggtitle("Scatter Plot of Red Wine pH and Quality Score.")+
  geom_smooth(method="lm", col="red")
```

Normality of the Residuals
``` {r 2.a2}
wine.lm<-lm(quality~pH, data=wine)
residual<-wine.lm$residuals
stand.residual<-rstandard(wine.lm)
xvalue<-wine$pH
yvalue<-wine$quality

diag.df<-data.frame(xvalue, yvalue, residual, stand.residual)

ggplot(diag.df, aes(sample=stand.residual))+
  stat_qq(col="blue")+
  stat_qq_line(col="red")+
  ggtitle("Normal Probability Plot of Red Wine Residuals")
```

Homoscedasticity - Equal Variance between Residuals
``` {r 2.a3}
ggplot(diag.df, aes(x=xvalue, y=residual))+
  geom_point(col="blue")+
  xlab("Predicted pH of Red Wines")+
  ylab("Residuals")+
  ggtitle("Plot pH Values of Red wines to Residuals")+
  geom_hline(yintercept=0, color="red", linetype="dashed")
```

#### Analysis
``` {r 2.b}
summary(wine.lm)
cor(wine$pH, wine$quality)
```

The estimated $\beta_1$ is -0.3020 with a p-value of 0.021. The adjusted $r^2$ value is 0.002709 and the correlation coefficient is -0.057. 

#### Conclusion
The p-value from the linear model is 0.021 which smaller than the alpha we set at 0.05. We reject the null hypothesis and conclude that there is a significant linear relationship between the pH value and the quality score of the red wines. 

However, despite the significant linear relationship, the low adjusted $r^2$ values and small correlation coefficient indicate that the negative linear relationship is very weak. 