---
title: "DATA602ProjectWineQ2Alt"
author: "Yu Hao"
date: "2025-10-06"
output: pdf_document
---

``` {r, message=FALSE, warning=FALSE}
rm(list=ls(all=TRUE))
library(ggplot2)
library(mosaic)
library(knitr)
opts_chunk$set(tidy.opts=list(width.cutoff=60),tidy=TRUE)
```

### Question 2: Is the pH level of the resultant *vinho verde* red wine significantly correlated with its perceieved quality?

Following the same notation as Question 1, low quality wine is defined as red wines with a quality score of 5 and less while high quality wine is defined as red wines with a quality score of 6 and more. 

Based off of the distribution of pH values in this sample of red wines the cutoffs of pH groups have been categorized in to <3.0, 3.0-3.5 and >3.5. 

#### Statistical Hypotheses
$H_0$: The quality of the red wine is indepedent from the pH level of the red wine. 

$H_A$: The quality of the red wine is not independent from the pH level of the red wine. 

#### Data Wrangling and Assumption Checking
``` {r 2.a1}
wine.df<-read.csv("C:\\Users\\yuhao\\Desktop\\Uni\\DATA 602\\winequality-red.csv")

pH.cat<-c()
for (i in wine.df$pH){
  if (i <= 3.0){
    pH.cat<-c(pH.cat, 1)
  }
  else if (i > 3.0 & i <= 3.5){
    pH.cat<-c(pH.cat, 2)
  }
  else {
    pH.cat<-c(pH.cat, 3)
  }
}

quality.cat<-c()
for (i in wine.df$quality){
  if (i <= 5){
    quality.cat<-c(quality.cat, 0)
  }
  else {
    quality.cat<-c(quality.cat, 1)
  }
}

wine.df<-data.frame(wine.df, pH.cat, quality.cat)
```
``` {r 2.a2}
crosstab.values<-c()
for (i in 0:1){
  for (j in 1:3){
    crosstab.values<-c(crosstab.values,length(subset(wine.df, quality.cat==i&pH.cat==j)$quality))
  }
}
crosstab.values

wine.crosstab<-rbind(c(18, 644, 82), c(17, 752, 86))
colnames(wine.crosstab)<-c("<3.0", "3.0-3.5", ">3.5")
rownames(wine.crosstab)<-c("Low", "High")
wine.crosstab
```
No category has a count of less than 5. 

#### Analysis
``` {r 2.b}
xchisq.test(wine.crosstab, correct=F)
```
$\chi^2$ statistic = 0.77742 and p-value = 0.6779.

#### Conclusion
The estimated p-value from this Chi-squared test of independence is 0.6779 which is larger than our set alpha of 0.05. We fail to reject to reject the null hypothesis and conclude that the quality of the red wines are not significantly dependent on their pH level. 
